XXX in a Tracing JIT Compiler for Efficient Dynamic Languages

### 2011-03-24 talk/icooolps2011/paper.tex
write background and start intro

XXX how exactly
the hints work that interpreter authors can use to improve the execution speed
of the programs running on their interpreters?

@@

PyPy's approach to implementing dynamic
languages is to write an interpreter for
the language in RPython. This interpreter can be translated to C and then
further to machine code. The interpreter consists of code in the form of a
large number of generated C functions and some data. Similarly, the user
program consists of functions in the language the interpreter executes.

XXX \cite{armin_rigo_pypys_2006}

@@

@@

>>>>

One of the hardest parts of implementing a dynamic language efficiently is to
optimize its object model. This is made harder by the fact that many recent
languages such as Python, JavaScript or Ruby have rather complex core object
semantics. For them, implementing just an interpreter is already an arduous
task. Implementing them efficiently with a just-in-time compiler is
nigh-impossible, because or their many corner-cases.

long dream of PE to not have to implement a compiler
new life by meta-tracers, spur, pypy, other things
trace execution of object model and get semantics right, not have to compile
object model

in spur and pypy, this can be improved by the interpreter author by adding
hints, which influence tracer and optimizer

pypy's hints go further than spurs, in this paper we present two very important
ones and show how classical implementation techniques of dynlangs can be
expressed by them

@@

The PyPy project \cite{armin_rigo_pypys_2006} strives to be an environment where
complex dynamic languages can be efficiently implemented. The approach taken
when implement a language with PyPy is to write an interpreter for the language
in \emph{RPython}. RPython is a restricted subset of Python chosen in such a way
that it is possible to perform type inference on it. The interpreters in RPython
can therefore be translated to efficient C code.

A number of languages have been implemented with PyPy, most importantly a full
Python implementation, but also a Prolog interpreter \cite{XXX} and a Smalltalk
VM \cite{XXX}.

This translation to C code adds a number of implementation details into the
final executable that are not present in the interpreter implementation, such as
a garbage collector. The interpreter can therefore be kept free from low-level
implementation details. Another aspect of the final VM that is added
semi-automatically to the generated VM is a tracing JIT compiler.

We call the code that runs on top of an interpreter implemented with PyPy the
\emph{user code} or \emph{user program}.

@@

XXX citations
A recently popular approach to JIT compilers is that of tracing JITs. Tracing
JITs record traces of concrete execution paths through the program. Those traces
are therefore linear list of operations, which are optimized and then get turned
into machine code. To be able to do this recording, VMs with a tracing JIT
typically also contain an interpreter. After a user program is started the
interpreter is used until the most important paths through the user program are
turned into machine code.

Because the traces always correspond to a concrete execution they cannot contain
any control flow splits. Therefore they encode the control flow decisions needed
to stay on the trace with the help of \emph{guards}. Those are operations that
check that the assumptions are still true when the trace is later executed with different values.

One disadvantage of tracing JITs which makes them not directly applicable to Pypy,

@@

XXX trace makes the object model operations explicit and transparent to the
optimizer
